### 2023-11-08 manuscript/manuscript.Rmd
Abbreviate ionisation products to IPs

Metabolite annotation is a crucial aspect for ensuring that biological interpretation is possible from metabolomics data sets.
A number of strategies can be employed to annotate peaks in mass spectra.
These can include simple approaches such as heuristic filtering of molecular formulas generated from *m/z* or direct matching putative ionisation products to metabolite databases [@draper2009; @green2015; @kind2007];
but also ensemble approaches that integrate a number of strategies like correlation analysis, ionisation product scoring and biotransformations [@baran2013; @fernandez2014; @kuhl2012; @uppal2017].
Tandem MS can also be employed for LC-MS approaches and provides a convenient method for structural validation [@neumann2010].
However, this is unsuitable for metabolome fingerprinting-based techniques due to the presence of structural isomers and isobaric compounds.
Many of the available software packages implementing these strategies are suitable for LC-MS techniques only.

@@

Features matching to possible ionisation products of the chemical standards were identified by first applying the MZedDB ionisation 'rules' and calculating their theoretical *m/z* for each of the adducts specified in Section \@ref(relationships) [@draper2009].
The calculated *m/z* were then matched to the *m/z* features in the standards mix and spiked urine data using a search range of $\pm$ `r search_ppm` ppm.

@@

To assess the performance of the molecular formula and graphical component selection approaches, the assignment outcomes were evaluated for the theoretical ionisation products calculated for the chemical standards that matched to *m/z* features in the standards mix and spiked urine. 
The standards mix allows for a direct evaluation of the selection approaches in a relatively simple, synthetic matrix.
Alternatively, the spiked urine provides an indication of how the selection approaches perform in a more realistic context where there are more *m/z* features present greater competition for assignment. 
To simplify interpretation, these evaluations were based on the assignment outcomes of the non-isotopic ionisation products during the adduct and isotopic assignment iterations.
This reduces the incidence of overlap between ionisation products, where multiple theoretical ionisation products match to a single *m/z* feature as well as a removes the inclusion of relationships involving transformations that complicates the calculation and interpretation of relevant relationships.
Based on the `r n_adducts` adduct rules specified, there were `r nrow(standards_assignment_outcomes_non_iso)` and `r nrow(spiked_urine_assignment_outcomes_non_iso)` theoretical ionisation products for these chemical standards that matched to `r n_features$standards` and `r n_features$spiked_urine` *m/z* in the standards mix and spiked urine respectively.
The theoretical ionisation products that directly matched *m/z* features, including the isotopic theoretical ionisation products, are provided in Additional file 5.
The frequencies of assignment outcomes for the theoretical ionisation products in these matrices are shown in Table \@ref(tab:ip-outcomes).

@@

Of the `r standards_n_AI_relevant` chemical standards that had theoretical ionisation products with relevant relationships in the standards mix, the molecular formula and graphical component selection approaches were able to correctly assign molecular formulas to `r standards_n_AI_assigned` (`r standards_percent_AI_assigned`%) of these.
For the `r xfun::numbers_to_words(standards_n_alternative)` ionisation products for which an alternative adduct and molecular formula were assigned, two of these were assigned to alternative components with a higher $P_c$ score resulting from a higher component AIS score.
The other alternative component was selected due to higher $P_c$ score resulting from a higher component degree.
There was only a single instance where an ionisation product was incorrectly assigned as a result of the molecular formula selection approach.

@@

The assignment results for the chemical standards in the standards mix and spiked urine are provided in Table \@ref(tab:compound-frequencies).
These show that, where compounds had *m/z* features matching to theoretical ionisation products, the approach was able to assign `r compounds_percent_assigned$standards`% and `r compounds_percent_assigned$spiked_urine`% of these chemical standards in the standards mix and spiked urine respectively. 
The *m/z* for which a chemical standard was assigned are provided in Additional file 6.

@@

While not the intended primary application of the approach, the C18-UPHLC-HRMS data from *B. distachyon* had the highest percentage of *m/z* features assigned along with the greatest number of molecular formulas. 
The chromatographic separation of the C18-UHPLC-HRMS enables the correlation network to be considered within limited retention time windows and increases the likelihood that a number of correlated ionisation products will be present for a given compound.
Added to this, the LC-MS reduces the effects of ion suppression compare to FIE-HRMS and also decreases the chance of potential interference in the correlations between related features due to the presence of structural isomers and isobaric compounds [@gowda2014].

@@

Using correlation networks of *m/z* feature abundances, molecular formula selection and graphical component selection, assigned molecular formulas for predicted ionisation products could be recovered from a subset of chemical standards spiked into an example human urine sample analysed using FIE-HRMS metabolome fingerprinting.
It was also possible to assign molecular formulas of common primary and secondary metabolites present in the KEGG compound database to *m/z* features from leaf tissue of *B. distachyon* using both FIE-HRMS fingerprinting and C18-UHPLC-HRMS profiling techniques.

>>>>

Metabolite annotation is a crucial aspect for ensuring that biological interpretation is possible from metabolomics data sets.
A number of strategies can be employed to annotate peaks in mass spectra.
These can include simple approaches such as heuristic filtering of molecular formulas generated from *m/z* or direct matching putative ionisation products (IPs) to metabolite databases [@draper2009; @green2015; @kind2007];
but also ensemble approaches that integrate a number of strategies like correlation analysis, ionisation product scoring and biotransformations [@baran2013; @fernandez2014; @kuhl2012; @uppal2017].
Tandem MS can also be employed for LC-MS approaches and provides a convenient method for structural validation [@neumann2010].
However, this is unsuitable for metabolome fingerprinting-based techniques due to the presence of structural isomers and isobaric compounds.
Many of the available software packages implementing these strategies are suitable for LC-MS techniques only.

@@

Features matching to possible IPs of the chemical standards were identified by first applying the MZedDB ionisation 'rules' and calculating their theoretical *m/z* for each of the adducts specified in Section \@ref(relationships) [@draper2009].
The calculated *m/z* were then matched to the *m/z* features in the standards mix and spiked urine data using a search range of $\pm$ `r search_ppm` ppm.

@@

To assess the performance of the molecular formula and graphical component selection approaches, the assignment outcomes were evaluated for the theoretical IPs calculated for the chemical standards that matched to *m/z* features in the standards mix and spiked urine. 
The standards mix allows for a direct evaluation of the selection approaches in a relatively simple, synthetic matrix.
Alternatively, the spiked urine provides an indication of how the selection approaches perform in a more realistic context where there are more *m/z* features present greater competition for assignment. 
To simplify interpretation, these evaluations were based on the assignment outcomes of the non-isotopic IPs during the adduct and isotopic assignment iterations.
This reduces the incidence of overlap between IPs, where multiple theoretical IPs match to a single *m/z* feature as well as a removes the inclusion of relationships involving transformations that complicates the calculation and interpretation of relevant relationships.
Based on the `r n_adducts` adduct rules specified, there were `r nrow(standards_assignment_outcomes_non_iso)` and `r nrow(spiked_urine_assignment_outcomes_non_iso)` theoretical IPs for these chemical standards that matched to `r n_features$standards` and `r n_features$spiked_urine` *m/z* in the standards mix and spiked urine respectively.
The theoretical IPs that directly matched *m/z* features, including the isotopic theoretical IPs, are provided in Additional file 5.
The frequencies of assignment outcomes for the theoretical IPs in these matrices are shown in Table \@ref(tab:ip-outcomes).

@@

Of the `r standards_n_AI_relevant` chemical standards that had theoretical IPs with relevant relationships in the standards mix, the molecular formula and graphical component selection approaches were able to correctly assign molecular formulas to `r standards_n_AI_assigned` (`r standards_percent_AI_assigned`%) of these.
For the `r xfun::numbers_to_words(standards_n_alternative)` IPs for which an alternative adduct and molecular formula were assigned, two of these were assigned to alternative components with a higher $P_c$ score resulting from a higher component AIS score.
The other alternative component was selected due to higher $P_c$ score resulting from a higher component degree.
There was only a single instance where an ionisation product was incorrectly assigned as a result of the molecular formula selection approach.

@@

The assignment results for the chemical standards in the standards mix and spiked urine are provided in Table \@ref(tab:compound-frequencies).
These show that, where compounds had *m/z* features matching to theoretical IPs, the approach was able to assign `r compounds_percent_assigned$standards`% and `r compounds_percent_assigned$spiked_urine`% of these chemical standards in the standards mix and spiked urine respectively. 
The *m/z* for which a chemical standard was assigned are provided in Additional file 6.

@@

While not the intended primary application of the approach, the C18-UPHLC-HRMS data from *B. distachyon* had the highest percentage of *m/z* features assigned along with the greatest number of molecular formulas. 
The chromatographic separation of the C18-UHPLC-HRMS enables the correlation network to be considered within limited retention time windows and increases the likelihood that a number of correlated IPs will be present for a given compound.
Added to this, the LC-MS reduces the effects of ion suppression compare to FIE-HRMS and also decreases the chance of potential interference in the correlations between related features due to the presence of structural isomers and isobaric compounds [@gowda2014].

@@

Using correlation networks of *m/z* feature abundances, molecular formula selection and graphical component selection, assigned molecular formulas for predicted IPs could be recovered from a subset of chemical standards spiked into an example human urine sample analysed using FIE-HRMS metabolome fingerprinting.
It was also possible to assign molecular formulas of common primary and secondary metabolites present in the KEGG compound database to *m/z* features from leaf tissue of *B. distachyon* using both FIE-HRMS fingerprinting and C18-UHPLC-HRMS profiling techniques.
